What Variables Impact Wicket Production in the IPL

### 2024-12-02 paper/paper.qmd
introduction first draft complete

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Siddharth Gowda
thanks: "Code and data are available at: [https://github.com/siddharthgowda/ipl_wicket_analysis](https://github.com/siddharthgowda/ipl_wicket_analysis)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

@@

## Estimand and Response Variable

The estimated for the research paper is likelihood a ball will result in a wicket. The goal of the paper is to understand when wickets occur. The wicket variable is the response variable.

@@

@@

# Model

The goal of the model is accurately predict when a wicket will occur in an [@ipl] game based on relevant variables from the dataset. Furtherore, I wanted to create a model that would achieve the most accuracy with the least amount of variables. To achieve this, three different generalized binomial family linear models (logistic regression) were used. The first model only had one input variable, the second one had two input variables, and third had four input variables. All models had the wicket boolean variable as the response variable.

@@

## Variables that Do Not Lead to Wickets

@@

@@

@@

```{r}
#| label: tbl-model-summary-four-var
#| tbl-cap: "Batting Style vs Bowling Style Do Not Heavily Impact Wicket Potential"
#| echo: false
#| warning: false
overly_complex_glm_wicket_model <- readRDS(here("models/overly_complex_glm_wicket_model.rds"))
#summary(overly_complex_glm_wicket_model)
modelsummary(overly_complex_glm_wicket_model)
```
Based on @tbl-model-summary-four-var, none of the batting styles or bowling style coefficient are statistically significant all with p values greater than 10%. This implies that most likely these variables not great predictors of wicket creation. The number of wickets in the previous over still has a statistically significant impact and has an increase of 0.031446 log odds per an increase of one wicket in the previous over. The residual deviance is 16574 on 51571 degrees of freedom compared to the null of 20849 on 51586 degrees of freedom.

## Variables Not Part of Model

```{r}
#| label: fig-stadium-wickets
#| fig-cap: Some stadiums are more prone to wickets than others.
#| echo: false
#| warning: false

stadium_boundaries <- cleaned_data %>%
  group_by(venue) %>%
  summarise(
    num_matches = n_distinct(match_id),
    num_wickets = sum(wicket == TRUE),  
  ) %>% arrange(desc(num_wickets), desc(num_matches))

ggplot(stadium_boundaries, aes(x = venue, y = (num_wickets/num_matches))) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(
       x = "Stadium Name",
       y = "Wickets Per Match") +
  theme_minimal() +
  coord_flip()

```
```{r}
#| label: fig-runrate-numofwickets
#| fig-cap: Current Run Rate Does Not Impact Wicket Rate
#| echo: false
#| warning: false

ggplot(cleaned_data, aes(x = run_rate, fill = factor(wicket))) +
  geom_histogram(position = "identity", alpha = 0.8, bins = 30) +
  labs(
    x = "Run Rate",
    y = "Number of Wickets",
  ) +
  scale_fill_discrete(name = "Wicket Occurred", labels = c("No Wicket", "Wicket")) + 
  theme_minimal()
```

>>>>

---
title: "What Variables Impact Wicket Production in the IPL"
subtitle: "Wicket can come in bunches in the IPL"
author: 
  - Siddharth Gowda
thanks: "Code and data are available at: [https://github.com/siddharthgowda/ipl_wicket_analysis](https://github.com/siddharthgowda/ipl_wicket_analysis)."
date: today
date-format: long
abstract: "The paper analyzes what variables are the best predictors of wickets in the Indian Permier League (IPL). The current over, the bowling style, batting hand, and the number of wickets in the previous over, impact the probability of a wicket occuring. Moreover, the number of wickets in the previous over the greatest predictor of wickets occuring. These findings can be used by IPL teams when attempting large comebacks or when trying to prevent comebacks."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

The game of cricket is simple. One player tries to bowl or delivery a ball to another player, the batsman, and the batsman tries to hit the ball onto the field in way that generates the most runs. Runs in cricket are scored by the two batsmen running between the wickets, the ball rolling out of the venue or if the ball is hit of the venue in the air. However, batters can get out or dismissed, and this called a wicket. In cricket, once a wicket occurs, a batsman cannot return to bat. As a result, one or two wickets can completely win change a game.

The goal of this paper is to figure out which variables are the best at predicting a future wicket in T20 Cricket. In T20 cricket, there are only 20 overs for each team to score. For more information on T20 cricket and cricket in generally read this [article](https://www.usatoday.com/story/sports/2024/06/01/what-is-out-over-wicket-cricket/73617682007/) [@usatoday]. The estimated for the research paper is the probability a ball will result in a wicket.

The analysis was done using IPL [@ipl] data using the `cricketdata` [@cricketdata] R [@citeR] CRAN package. All code and data analysis was created using R [@citeR]. The `tidyverse` [@tidyverse] package was used for data cleaning and feature creation, `knitr` [@knitr] was used for formatting tables, `here` [@here] was used for reading in data in a `parquet` format, `modelsummary` [@modelsummary] was used to generate summaries for the models created, and `marginaleffects` [@marginaleffects] was used to generate predictions based on the models.

The paper explore the IPL [@ipl] play-by-play from the 2021, 2022, 2023, and 2024 seasons. The paper first used descriptive statistical analysis techniques to figure out which variables most associated with wicket prediction. This incorporated summary tables, scatter plots with trend lines, histograms, and more. From this analysis, the current over, the number of wickets in the previous over, the bowling style, and the batting style were determined to be the most important factors in terms of wicket production.

From those variables, three logistic models were created ranging from lower complexity to high complexity. Lower complexity models had less input variables while high complexity models more input variables. The response variable was wicket occurrence, a Boolean indicating whether or not a wicket occurred. The training and testing split for these models was 80% to 20%. The model that used the over number and the number of wickets in the previous over as the input variables did the best job of predicting wickets. Ultimately, the number of wickets in the previous over was the best predictor of a wicket, suggesting that wickets in the IPL [@ipl] can come in bunches.

@@

@@

```{r}
#| label: fig-runrate-numofwickets
#| fig-cap: Current Run Rate Does Not Impact Wicket Rate
#| echo: false
#| warning: false

ggplot(cleaned_data, aes(x = run_rate, fill = factor(wicket))) +
  geom_histogram(position = "identity", alpha = 0.8, binwidth = 1) +
  labs(
    x = "Run Rate",
    y = "Number of Wickets",
  ) +
  scale_fill_discrete(name = "Wicket Occurred", labels = c("No Wicket", "Wicket")) + 
  theme_minimal()
```

From @fig-runrate-numofwickets, both the distribution of current run rates of teams when a wicket occurs and when a wicket does not occur are approximately normally distributed around 8.5. The variability for both distributions also appears to be the same. However, the distribution of run have no wickets occur has more outliers than when wickets occur.

@@

# Model {#sec-model}

The goal of the model is accurately predict when a wicket will occur in an [@ipl] game based on relevant variables from the dataset. Furthermore, I wanted to create a model that would achieve the most accuracy with the least amount of variables. To achieve this, three different generalized binomial family linear models (logistic regression) were used. The first model only had one input variable, the second one had two input variables, and third had four input variables. All models had the wicket Boolean variable as the response variable.

@@

From the analysis, there were three variables that seemed to impact the probability of a wicket occurring: the over the ball was bowled in, the number of wickets in the previous over, and the bowling style and the batting style match up. From @fig-onevarmodel-results and @fig-over-wickets, there is a strong evidence to suggest that wickets are more likely to happen later in overs. This is likely true for two reasons. One is that in T20 cricket, batters are willing to take more risks in later over because there are only a few balls left and they want to maximize the amount of runs they can score. The other interpret is that in later over, top batsman are usually already out, so lower quality batsman are playing and they are more likely to get out. As a result, wicket occurrences are more likely. While this finding is not anything most cricket players or coaches do not know, it is still important to see it backed up through data since oftentimes convention wisdom is not actually true. Also, from @fig-over-wickets, the fifth over seems to have an unusually high amounts of wickets than the trend line would suggest. However, this makes sense since the fifth over is the end of the power play, where fielders have to be positioned closer to the pitch making it easier to score runs. Therefore, batsman might be willing to take more risks in the 5th over so they can maximize the runs scored in the power play, leading to more wickets.

Based on @fig-twovarmodel-results and @fig-prevoverwickets-wickets, if more wickets were taken in the previous over, the more likely it is to get a wicket in the next over. This also can be interpreted in two ways. One is that when more wickets are taken in the previous over, worse batsman are playing in the next over are worse batsman are more likely to get out than better batsman. The other interpretation is that when a lot of wickets occur in the previous over, new batsman come in and new batsman are not in rhythm and are still trying to figure out what the bowlers' strategies. Thus, they are more likely to get out, resulting in a wicket. From the econ interpretation, if a team is behind by a lot in a match and suddenly gets a wicket, in the next over, the captain could go to their top wicket producing bowler in order to get more wickets to put them back in the game.

Via @fig-bowlingstyle-wickets, certain bowling and batting match-ups result in more wickets. The general trend is that fast bowlers are more likely to take wickets compared to spinners (wrist spin, leg break, leg break googly, off break). The chart showed that left hand fast bowlers had the highest wicket taking percentage against left handed batters. and that right hand medium and right hand medium fast bowlers had the highest wicket taking percentage against right handed batters. However, there are limitations to these conclusions. First, there were only four left-hand fast bowlers in this dataset (@tbl-bowling-style), and certain there are a lot more right handed batsman than left handed batsman (@tbl-batting-style). However, assumption the trends in the graph are true, a captain could potentially use to choose bowlers that give them the best chance of a wicket, depending on the match up.

From @fig-runrate-numofwickets, there does not appear to be a relationship between the current run rate of a team and if a wicket occurs. This is important because it shows that teams should not merely assume that if the opposition has an extremely high run rate, they are taking too many risks and will eventually get out.

@@

From @sec-model, three models were created, all predicting the probability a wicket would occur. The first model used the over as the only input; the second model the over and the number of previous wickets in the over as the two input variables; and the third model used the second model the over, the number of previous wickets in the over as the two input variables, the bowling style, and the batting style (left hand or right hand). It is clear that the second model is the most effective at predicting when a wicket will occur. While the first model does technically predict when a wicket occurs more accurately than the second model (95.000% compared to 94.952%), the first model never correctly predicting when a wicket would occur in the test data (@tbl-model-one-var-results). In terms of the third model, none of the variables' coefficients were statistically significant except for the number of wickets in the previous over. Therefore, that model cannot be the best. However, while the second model is the best predictor, it also rarely predicted when a wicket would occur (2.640% wicket prediction accuracy).

Based on the second model, it clear that the number of wickets in the previous over is by far the most important variables when predicting wicket occurrences. From @fig-twovarmodel-results from a jump from zero to two wicket in the previous over almost guaranteed a wicket prediction from the model. This is important because it demonstrates than in the IPL wickets can potentially come in bunches. Therefore, if a team is leading by a significant margin in a match and one of their batters gets out, the captain might be wise to choose a more stable batsman to replace him rather than one that might score more runs, but is more likely to get out.

@@

The biggest limitation of this study comes from the fact that only IPL [@ipl] data was used. The IPL  [@ipl] is a T20 Cricket League that has the best players in the world, so it will be difficult to use these findings in other leagues and especially in different cricket formats like test and one-day-international. The IPL [@ipl] pitches and stadiums generally favor offense and scoring a lot of runs compared to bowling. This also makes the conclusions from this paper less applicable in other cricket leagues.

The biggest weakness of this paper comes from the lack of fielding data. In the future, the paper would incorporate play-by-play data that shows the position of the fielders and who recovers a ball and some distance and speed data, like distance between ball and nearest fielder and speed or rpms (revolutions per minute) of a bowling delivery. Also in the future, the paper would incorporate data from other T20 cricket leagues to verify if these findings can be applicable to all form of T20 cricket.

@@

```{r}
#| label: tbl-model-summary-four-var
#| tbl-cap: "Batting Style vs Bowling Style Do Not Heavily Impact Wicket Probability"
#| echo: false
#| warning: false
overly_complex_glm_wicket_model <- readRDS(here("models/overly_complex_glm_wicket_model.rds"))
#summary(overly_complex_glm_wicket_model)
modelsummary(overly_complex_glm_wicket_model)
```
Based on @tbl-model-summary-four-var, none of the batting styles or bowling style coefficient are statistically significant all with p values greater than 10%. This implies that most likely these variables not great predictors of wicket creation. The number of wickets in the previous over still has a statistically significant impact and has an increase of 0.031446 log odds per an increase of one wicket in the previous over. The residual deviance is 16574 on 51571 degrees of freedom compared to the null of 20849 on 51586 degrees of freedom.
